---
title: "Lipidomics"
author: "Nikeisha Caruana"
date: "21/08/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
pacman::p_load("stringr","gplots","ggplot2","ComplexHeatmap", "limma", "edgeR", "Homo.sapiens", "dplyr", "tidyr", "readr", "circlize", "RColorBrewer", "dendextend", "imputeLCMD", "impute")
```

Read in data
```{r}
TotalLipids <- read.csv("data/2020_08_12_MP20_007_final_totals.csv")
CL <- read.csv("data/2020_04_06_MP20_007_CL_total.csv")
CL <- na.omit(CL)
```

Clean Up
```{r}
#remove 0s columns
TotalLipids <- TotalLipids[, colSums(TotalLipids != 0) > 0]
CLsp <- CL[c(4:32)]
TotalLipids <- cbind(TotalLipids, CLsp)
rownames(TotalLipids) <- TotalLipids$sample.type
totallipidsmatrix <- TotalLipids[-c(1:3)]
totallipidsmatrix <- totallipidsmatrix[-c(2,3,4,6,9,11,12,13,14,21,34)]
totallipidsmatrix <- log2(totallipidsmatrix)
```

Just CLs for checking normalisation
```{r}
allCLs <- CL[c(4:32)]

#remove 0s
allCLs <- na_if(allCLs, 0)
allCLs <- allCLs[ , colSums(is.na(allCLs)) == 0]
rownames(allCLs) <- CL$Sample.type 
```

Add Total CL to other Totals
```{r}
totalsall <- cbind(TotalLipids, totalcl = CL$Total.CL)
rownames(totalsall) <- totalsall$sample.type
totalsdata <- totalsall[c(4:43)]

#remove 0s
alllipids <- na_if(totalsdata, 0)
alllipids <- alllipids[ , colSums(is.na(alllipids)) == 0]
```

Sample Annotation
```{r}
sampleanno <- as.data.frame(TotalLipids$sample.type) %>% separate('TotalLipids$sample.type', c("subject", "trainingvolume"), sep = '-')
sampleanno <- sampleanno %>% 
  mutate(trainingvolume = str_replace(trainingvolume, "1", "BL")) %>% 
  mutate(trainingvolume = str_replace(trainingvolume, "2", "NVT")) %>% 
  mutate(trainingvolume = str_replace(trainingvolume, "5", "HVT")) %>% 
  mutate(trainingvolume = str_replace(trainingvolume, "8", "RVT"))
sampleanno <- cbind(sampleanno, totalsall$sample.type)
rownames(sampleanno) <- sampleanno$`totalsall$sample.type`
```

~~CARDIOLIPINS~~

Normalisation CLs
```{r cardiolipins}
#normalise the data - take the mean of the log 
lipid_logCL <- log2(allCLs)
lipid_normCL <- normalizeQuantiles(lipid_logCL)
rownames(lipid_logCL) <- sampleanno$`totalsall$sample.type`
```

MDS plot CL
```{r}
#check mds
T_norm <- t(lipid_logCL)
mds_data <- plotMDS(T_norm, plot = TRUE)
mds_data <- data.frame(x=mds_data$x,y=mds_data$y,sampleanno[names(mds_data$x),])

p <- ggplot(mds_data,aes(x=x,y=y)) + geom_point(aes(color = trainingvolume), size=6)
p
```

Differential Expression analysis 
```{r}
#DE
design_all <- model.matrix(~0 + trainingvolume, sampleanno)
colnames(design_all) <- gsub("trainingvolume","", colnames(design_all))
rownames(design_all) <- sampleanno$sample

contr.matrix_musclelipids <- makeContrasts(
    NTVvsBase = NVT-BL,
    HTVvsBase = HVT-BL,
    RTVvsBase = RVT-BL,
    HTVvsNTV = HVT-NVT,
    RTVvsNTV = RVT-NVT,
    HTVvsRTV = RVT-HVT,
    levels = design_all
  )

all_fit <- lmFit(T_norm, design_all)
cfit_all <- contrasts.fit(all_fit, contrasts = contr.matrix_musclelipids)
all_efit<-eBayes(cfit_all)

all_comp_lips <- topTable(all_efit, coef=c(1,4,6), n=Inf, adjust.method = "BH")
all_comp_lips2 <- subset(all_comp_lips, all_comp_lips$adj.P.Val<0.05)
```

Profile Plots - shows CLs increase overall with exercise - therefore need to normalise the same way as proteomics. 

```{r}
mergeCLs <- cbind(lipid_logCL, sampleanno)
mergeCLs$total <- rowSums(mergeCLs[,c(1:25)])

mergeCLs$trainingvolume <- factor(mergeCLs$trainingvolume, c("BL", "NVT","HVT","RVT"))

#pdf("profileplot_rawprotein.pdf", height = 15, width = 20)
ggplot(data = mergeCLs, aes(x = trainingvolume, y = total, group = subject, color = subject)) + geom_smooth(se = FALSE) 

avgCL <- mergeCLs[c(26:29)]
avgCL<- avgCL %>%
  group_by(trainingvolume) %>%
  summarise(mean = mean(total), n = n())
ggplot(data = avgCL, aes(x = trainingvolume, y = mean, group = 1)) + geom_line()

```

```{r}
averageCLs <- as.data.frame((mergeCLs[c(1:25)]))
averageCLs <- cbind(averageCLs, mergeCLs[c(26,27)])

averageCLs <- averageCLs %>% gather("lipid","concentration", -subject, -trainingvolume)
averageCLs <- averageCLs  %>% group_by(trainingvolume)
averageCLs$trainingvolume <- factor(averageCLs$trainingvolume, c("BL","NVT","HVT","RVT"))

averageCLmean <- averageCLs %>%
  dplyr::group_by(trainingvolume) %>%
   summarise(mean = mean(concentration), n = n())

#pdf("profileplot_allCL_unnormed.pdf", height = 15, width = 18)
ggplot(data = averageCLmean, aes( x =trainingvolume, y = mean, group = 1)) + geom_line()
#dev.off()
```

check CLs individual species to make sure they go up. 
```{r}

#subset the CLs first
OGCls <- mergeCLs[-c(29)]

OGCls_long <- OGCls %>% gather("Species","intensity", -subject, -trainingvolume, -`totalsall$sample.type`)
OGCLscaled <- OGCls_long %>% group_by(subject, Species) %>% mutate(z_score = scale(intensity))

OGCLprofileplot_mean <- OGCLscaled %>% group_by(Species, trainingvolume) %>% dplyr::mutate(Mean = mean(z_score, na.rm=TRUE)) 

#pdf("profileplot_CLOG_20210112.pdf", height = 6, width = 7)
ggplot(data = OGCLprofileplot_mean, aes( x = trainingvolume, y = Mean, group = Species)) + geom_smooth() + theme_bw() + theme(text = element_text(size=20))
#dev.off()
```

LIPID DATA NORMALISED BY TOTAL CLS


SPECIES NORMALISED BY TOTAL CL

```{r SPECIES NORMALISED BY TOTAL CL - THIS IS BETTER}
CLnormed_sp <- read.csv("data/2020_04_06_MP20_007_CL_normedspecies.csv", na.strings = "na", check.names = FALSE)
CLnormed_sp <- CLnormed_sp[-c(1),]
rownames(CLnormed_sp) <- CLnormed_sp$'sample type'
CLnormed_sp <- CLnormed_sp[-c(1:3)]
CLnormed_sp <- t(CLnormed_sp)
CLnormed_sp <- CLnormed_sp[!rowSums(is.na(CLnormed_sp)) > ncol(CLnormed_sp)*.3,]


class(CLnormed_sp) <- "numeric"
```

ALL CL NORMED LIPIDS
```{r}
batch <- sampleanno$subject
removedbatch <- removeBatchEffect(CLnormed_sp, batch = batch)

#check mds
mds_data <- plotMDS(removedbatch, plot = TRUE)
mds_data <- data.frame(x=mds_data$x,y=mds_data$y,sampleanno[names(mds_data$x),])
p <- ggplot(mds_data,aes(x=x,y=y)) + geom_point(aes(color = trainingvolume), size=3)
p
```

```{r}
imputedsp <- impute.knn(CLnormed_sp)
imputedsp <- imputedsp[[1]]
```

Normalisation
```{r}
#normalise the data - take the mean of the log 
clnorm_logsp <- log2(imputedsp)
```

zscore check
```{r}
longsp <- as_tibble(clnorm_logsp)
lipidnames <- rownames(clnorm_logsp)
longsp$lipidnames<- lipidnames
longsp2 <- longsp %>% gather("sample","intensity", -lipidnames)
longsp2 <- longsp2 %>% separate(sample, c("subject", "trainingvolume"))
scaledsps <- longsp2 %>% group_by(subject,lipidnames) %>% mutate(z_score = scale(intensity))
scaledsps <- scaledsps[,-4]
scaledsps <- scaledsps %>% unite("sample", subject, trainingvolume)
widesp <- scaledsps %>% spread(sample,z_score)
```

Profile Plots - Z-Scored
```{r}
zscore_grouped <- read.table("data/zscorespecies_class2.csv", sep = ",", check.names = FALSE, header=TRUE)
groupedinfo <- zscore_grouped[c(1:3)]
widesp <- as.data.frame(widesp)
zscore_grouped <- cbind(groupedinfo, widesp)

zgroup_long <- zscore_grouped %>% gather("Volume","intensity", -Class, -Species, -Group, -lipidnames)
zgroup_long <- zgroup_long %>% separate(Volume, c("subject", "trainingvolume"))

zgroup_long_profile <- zgroup_long  %>% group_by(trainingvolume)
zgroup_long_profile$trainingvolume <- factor(zgroup_long_profile$trainingvolume, c("1","2","5","8"))
zgroup_long_profile$intensity <- as.numeric(zgroup_long_profile$intensity)

#zgroup_long <- zgroup_long[-c(1)]
zgroupmean <- zgroup_long %>%
  dplyr::group_by(Class,Group,trainingvolume) %>%
   summarise(mean = mean(intensity), n = n())

ggplot(data = zgroupmean, aes( x =trainingvolume, y = mean, group = Class, color = Group)) + geom_smooth() + facet_wrap(Class ~ .) + theme_bw() + theme(text = element_text(size=20))
```

cardiolipin profile plot normalised, all CLs 
```{r}
#subset the CLs first
CLprofileplot_all <- zgroup_long %>% filter(Class == "CL")

CLprofileplot_mean <- CLprofileplot_all %>% group_by(Species, trainingvolume) %>% dplyr::mutate(Mean = mean(intensity, na.rm=TRUE)) 

ggplot(data = CLprofileplot_mean, aes( x = trainingvolume, y = Mean, group = Species)) + geom_smooth() + theme_bw() + theme(text = element_text(size=20))
```

Differential Expression analysis 
```{r}
widesp <- as.data.frame(widesp)
rownames(widesp) <-widesp$lipidnames
widesp2 <- widesp[-c(1)]

#DE
design_all <- model.matrix(~0 + trainingvolume + subject, sampleanno)
colnames(design_all) <- gsub("trainingvolume","", colnames(design_all))
#colnames(design_all) <- gsub("subject","", colnames(design_all))
rownames(design_all) <- sampleanno$`totalsall$sample.type`

contr.matrix_musclelipids <- makeContrasts(
    NTVvsBase = NVT-BL,
    HTVvsBase = HVT-BL,
    RTVvsBase = RVT-BL,
    HTVvsNTV = HVT-NVT,
    RTVvsNTV = RVT-NVT,
    HTVvsRTV = RVT-HVT,
    levels = design_all
  )

all_fit <- lmFit(widesp2, design_all)
cfit_all <- contrasts.fit(all_fit, contrasts = contr.matrix_musclelipids)
all_efit<-eBayes(cfit_all)

all_comp_lips <- topTable(all_efit, c(1,4,6), n=Inf, adjust.method = "BH",confint=TRUE)
all_comp_lips2 <- subset(all_comp_lips, all_comp_lips$adj.P.Val<0.05)
all_comp_lips2$names <- rownames(all_comp_lips2)
```

Add Lipid group to assess results
```{r}
#heatmap annotation
groupnames <- zscore_grouped[c(1:3)]
groupnames <- cbind(groupnames, rownames(widesp2))

annot_DE <- merge(groupnames, all_comp_lips2, by.x="rownames(widesp2)", by.y = "names")
```

```{r}
lipidanno <- HeatmapAnnotation(lipid=annot_DE$Group, which="row", annotation_width=unit(2, "cm"), gap=unit(1, "mm"), simple_anno_size = unit(2, "cm"))

col = colorRamp2(c(-1.5, 0, 1.5), c("#272775", "white", "#B74E25"))

dendrow <- annot_DE[c(5:7)] %>% dist(method = "euclidean") %>% hclust(method = "complete") %>% as.dendrogram %>% color_branches(k = 6)

Heatmap(as.matrix(annot_DE[c(5:7)]), cluster_rows = dendrow, cluster_columns = FALSE,right_annotation = lipidanno,  col = col, rect_gp = gpar(col = "slategray", lwd = 0.1))
```
